A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-07 latex/main.tex
Updates from Overleaf

\author[1,7,*,\orcidlink{0000-0002-6351-8873}]{Shing~H.~Zhan}
\author[1,6,\orcidlink{0000-0002-3536-6411}]{Yan~Wong}
\author[2,6,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
\author[3]{Katherine~Eaton}
\author[1]{Isobel~Guthrie}
\author[1,\orcidlink{0000-0002-1982-6801}]{Benjamin~Jeffery}
\author[1,2,4,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
\author[3,\orcidlink{0000-0002-1543-4501}]{Carmen~Lia~Murall}
\author[5,\orcidlink{0000-0003-3042-0818}]{Sarah~P.~Otto}
\author[1,**,\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}

@@

% The ARG is big, took a decent chunk of time to infer (but not crazy)
% and is very easy to analyse.
Using sc2ts, we built an ARG from 2,482,157 samples
from the Viridian v04 dataset~\cite{Hunt2024},
collected between 2020-01-01 and 2023-02-20.
Inference took 21 days on a server with
64 cores (2$\times$ AMD EPYC 7502) and 512 GB of RAM
($\sim$5 CPU years;
Figures~\ref{fig:inference_resources}~\ref{fig:samples_per_day}).
After post-processing, the ARG contains
2,689,054 nodes,  % Note the estimated dates ?
2,748,838 edges, and
2,285,344 mutations
at 29,893 sites,
and is stored in a 32 MB file.
Of the nodes, 855 represent recombination events,
resulting in 316 distinct trees along the genome
(some events occurred at the same breakpoints).
Using the tskit Python API,
% NOTE: numbers from Yan's laptop are:
% ts = tszip.load(): user 484 ms, sys: 90.4 ms, total: 575 ms
% sc2ts.node_data(ts, inheritance_stats=False): user 396 ms, sys: 42.8 ms, total: 439 ms
the ARG can be loaded into memory in half a second,
and the vectorised representation integrates tightly
with powerful data science tools, such as NumPy~\cite{Harris2020array}.
For example, a tabular summary for the 2.75 million nodes,
including sample IDs and Pango assignments (stored as metadata in the ARG),
can be derived as a Pandas data frame~\cite{Mckinney2010data} in less than half a second.
The ARG can therefore be easily accessed using familiar tools
with minimal hardware requirements, enabling diverse bespoke analyses.

@@

\section*{DISCUSSION}
There are two primary contributions of this work. The first is an
ARG for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analyses widely used in population genomics.
This combination of a deeply validated pandemic-scale ARG with 
a mature and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of ARG-aware methods in human 
genomics~\cite{Mejia2025using,Grundler2025geographic,
Lehmann2025args,Terhorst2025accelerated}.
We hope that the ARG provided here will similarly
enable a range of recombination-aware analyses of SARS-CoV-2, and 
catalyse a wave of methodological developments integrating
recombination into pathogen genomics.
We envisage the ARG as a long-term stable resource---building on the 
high-quality, freely available reassemblies of
the Viridian dataset~\cite{Hunt2024}---to facilitate the study 
of the pandemic phase of SARS-CoV-2 in an open and reproducible manner.
% NOTE: still missing the "What" bit suggested by Jotun, where we 
% list out some of the stuff that could be done. I'm not convinced
% this works, though.

@@

SHZ is supported by an NDPH Intermediate Research Fellowship
(Oxford Population Health, University of Oxford). 
% "acknowledges support" here means "was awarded this grant 
% and it supported this research"
SO acknowledges support from the Natural Sciences and Engineering Research Council of Canada (RGPIN-2022-03726).
JK acknowledges support from R01 HG012473 from the National Institutes of Health NHGRI, EPSRC (research grant EP/X024881/1) and the Robertson Foundation. DSP acknowledges funding administered by the Danish National Research Foundation in support of the Pioneer Centre for SMARTbiomed. For the purpose of Open Access, the authors have applied a CC BY public copyright licence to any Author Accepted Manuscript version arising from this submission.

>>>>

\author[1,7,*,\orcidlink{0000-0002-6351-8873}]{Shing~H.~Zhan}
\author[1,6,\orcidlink{0000-0002-3536-6411}]{Yan~Wong}
\author[2,6,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
\author[3]{Katherine~Eaton}
\author[1,\orcidlink{0000-0002-3547-7226}]{Isobel~Guthrie}
\author[1,\orcidlink{0000-0002-1982-6801}]{Benjamin~Jeffery}
\author[1,2,4,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
\author[3,\orcidlink{0000-0002-1543-4501}]{Carmen~Lia~Murall}
\author[5,\orcidlink{0000-0003-3042-0818}]{Sarah~P.~Otto}
\author[1,**,\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}

@@

% The ARG is big, took a decent chunk of time to infer (but not crazy)
% and is very easy to analyse.
Using sc2ts, we built an ARG from 2,482,157 samples
from the Viridian v04 dataset~\cite{Hunt2024},
collected between 2020-01-01 and 2023-02-20.
Inference took 21 days on a server with
64 cores (2$\times$ AMD EPYC 7502) and 512 GB of RAM
($\sim$5 CPU years;
Figures~\ref{fig:inference_resources}~\ref{fig:samples_per_day}).
After post-processing, the ARG contains
2,689,054 nodes,  % Note the estimated dates ?
2,748,838 edges, and
2,285,344 mutations
at 29,893 sites,
and is stored in a 32 MiB file.
Of the nodes, 855 represent recombination events,
resulting in 316 distinct trees along the genome
(some events occurred at the same breakpoints).
Using the tskit Python API,
% NOTE: numbers from Yan's laptop are:
% ts = tszip.load(): user 484 ms, sys: 90.4 ms, total: 575 ms
% sc2ts.node_data(ts, inheritance_stats=False): user 396 ms, sys: 42.8 ms, total: 439 ms
the ARG can be loaded into memory in half a second,
and the vectorised representation integrates tightly
with powerful data science tools, such as NumPy~\cite{Harris2020array}.
For example, a tabular summary for the 2.75 million nodes,
including sample IDs and Pango assignments (stored as metadata in the ARG),
can be derived as a Pandas data frame~\cite{Mckinney2010data} in less than half a second.
The ARG can therefore be easily accessed using familiar tools
with minimal hardware requirements, enabling diverse bespoke analyses.

@@

\section*{DISCUSSION}
There are two primary contributions of this work. The first is an
ARG for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analyses widely used in population genomics.
This combination of a deeply validated pandemic-scale ARG with 
a mature and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of ARG-aware methods in human 
genomics~\cite{Mejia2025using,Grundler2025geographic,
Lehmann2025args,Terhorst2025accelerated}.
We hope that the ARG provided here will similarly
enable a range of recombination-aware analyses of SARS-CoV-2, and 
catalyse a wave of methodological developments integrating
recombination into pathogen genomics.
For instance, annotating the ARG with experimentally derived
fitness estimates for mutations can help elucidate selection
pressures on different recombinants;
analysing the genomic positions of mutations occurring on long
branches (which persist after accounting for recombination) can
illuminate the evolution of saltational variants.
We envisage the ARG as a long-term stable resource---building on the 
high-quality, freely available reassemblies of
the Viridian dataset~\cite{Hunt2024}---to facilitate the study 
of the pandemic phase of SARS-CoV-2 in an open and reproducible manner.
% NOTE: still missing the "What" bit suggested by Jotun, where we 
% list out some of the stuff that could be done. I'm not convinced
% this works, though.

@@

SHZ is supported by an NDPH Intermediate Research Fellowship
(Oxford Population Health, University of Oxford). 
% JK: Commenting this out until exact wording is provided. I'm pretty
% sure that "Wellcome trust" is wrong.
% IG is supported by the Wellcome trust (DPhil in Genomic Medicine and Statistics).
DSP acknowledges funding administered by the Danish National Research Foundation in support of the Pioneer Centre for SMARTbiomed. 
% "acknowledges support" here means "was awarded this grant 
% and it supported this research"
SO acknowledges support from the Natural Sciences and Engineering Research Council of Canada (RGPIN-2022-03726).
JK acknowledges support from R01 HG012473 from the National Institutes of Health NHGRI, EPSRC (research grant EP/X024881/1) and the Robertson Foundation. For the purpose of Open Access, the authors have applied a CC BY public copyright licence to any Author Accepted Manuscript version arising from this submission.

Computation used the Oxford Biomedical Research Computing (BMRC) facility, a joint development between the Wellcome Centre for Human Genetics and the Big Data Institute supported by Health Data Research UK and the NIHR Oxford Biomedical Research Centre. The views expressed are those of the author(s) and not necessarily those of the NHS, the NIHR or the Department of Health.
